

title: "ARIMA NOTE"
author: "Joseph Rickert"
date: 2025-02-17
code-fold: true
code-summary: "Show the code"
description: "TimeGPT is a pre-trained, multi-layer, encoder/decoder transformer model with self-attention mechansims designed specifically for time series forecasting. This post, a revision of the of the post first published on 2025-02-12, corrects an error that deleteriously affected the ARIMA and exponential smoothing forecasts which are contrasted with the TimeGPT forecast."
categories: ""
editor: source
---



```{r}
#| message: FALSE
#| warning: FALSE
#| code-fold: true
#| code-summary: "Show the code"
library(tidyverse)
library(forecast)
library(fable)
library(tsibble)
library(feasts)
```


```{r}
#| code-fold: true
#| code-summary: "Show the code"
df <- nixtlar::electricity
glimpse(df)
```


```{r}
df2 <- df |> mutate(time = as.POSIXct(ds, format = "%Y-%m-%d %H:%M:%S")) |> 
             filter(unique_id == "BE") |> select(-unique_id, -ds)
  

p <- df2 |> ggplot(aes(x = time, y = y)) +
  geom_line(color='darkblue') +
  ggtitle(" BE Electricity Usage Data")

p
```

## Prepare Training and Test Data
```{r}
#| code-fold: true
#| code-summary: "Show the code"
NF <- 24

BE_df_wide <- df |> pivot_wider(names_from = unique_id, values_from = y) |>
  select(ds, BE) |> drop_na()

BE_train_df <- BE_df_wide %>% filter(row_number() <= n() - NF)
BE_test_df <- tail(BE_df_wide, NF)
BE_train_df <- BE_train_df |> rename(y = BE) |> mutate(unique_id = "BE")
BE_test_df <- BE_test_df |> rename(y = BE)
```




This next section of code reformat the training data into a format that is suitable for the `auto.arima()` which requires that the data be expressed as a `ts()` object. 

```{r}
#| code-fold: true
#| code-summary: "Show the code"


train <- BE_train_df |> select(-unique_id) |>
mutate(time = 1:length(ds))|> select(-ds)

elec_ts <- ts(train$y, frequency = 24)

```

Forecast 
```{r}
#| code-fold: TRUE
#| code-summary: "Show the code"
#| message: FALSE
#| warning: FALSE
#arima_1 <- elec_ts |>
#auto.arima() |>
#forecast(h = NF , level = 95)
#saveRDS(arima, "arima_1_24.rds")
arima <- readRDS("arima_1_24.rds")
```




```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| message: FALSE
#| warning: FALSE
arima_fcst_df <- BE_test_df |> 
  mutate(time = ds,
    BE_actual = y,
    arima = as.vector(arima$mean)) |> select(-ds,-y)
head(arima_fcst_df,3)

```




### fable



```{r}
auto_train <- BE_train_df |> select(-unique_id) |>
mutate(time = as.POSIXct(ds, format = "%Y-%m-%d %H:%M:%S")) |> select(-ds)
  
elec_ts_2 <- auto_train |> as_tsibble(index = time) |> fill_gaps(time, .full = start())
  
```




```{r}
fit <- elec_ts_2 %>%
  model(
    arima = ARIMA(y)
  )
fable_ARIMA_fcst <- fit |> forecast(h = 24)
fit
```


```{r}


#| code-fold: true
#| code-summary: "Show the code"

#ARIMA_2 <- as.vector(fable_ARIMA_fcst$.mean)
arima_fcst_df <- arima_fcst_df |> mutate(ARIMA =  as.vector(fable_ARIMA_fcst$.mean) )
head(arima_fcst_df,3)
```


```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| classes: preview-image
arima_fcst_long_df <- arima_fcst_df %>%
  pivot_longer(!time, names_to = "method", values_to = "mean")

q <- arima_fcst_long_df |>
  ggplot(aes(
    x = time,
    y = mean,
    group = method,
    color = method
  )) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  geom_line() +
  geom_point() +
  ggtitle("forecast vs. fable auto ARIMA forecasts")

q
```




```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| warning: false
RMSE <-  function(m, o){sqrt(mean((m - o)^2))}
rms_names <- c("arima", "ARIMA")
rms_fcst <- array(NA_real_,
                          dim = 2,
                          dimnames = list(rms_names))
rms_fcst[1] <- RMSE(arima_fcst_df$arima, arima_fcst_df$BE_actual)
rms_fcst[2] <- RMSE(arima_fcst_df$ARIMA, arima_fcst_df$BE_actual)


rms_fcst
```





## ARIMA(0,1,4)(0,0,2)[24]

This notation translates into:

$$(1 - B)y_t = (1 - \theta_1B - \theta_2 B^2 - \theta_3 B^3 - \theta_4 B^4)(1 - \Theta_1 B^{24} - \Theta_2 B^{48})\varepsilon_t$$
which expands into:



$$y_t−y_{t−1}=\varepsilon_t−\theta_1\varepsilon_{t−1}−\theta_2\varepsilon_{t−2}−\theta_3\varepsilon_{t−3}−\theta_4\varepsilon_{t−4}−\Theta_1\varepsilon_{t−24}+\theta_1\Theta_1\varepsilon_{t−25}+\theta_2\Theta_1\varepsilon_{t−26}+\theta_3\Theta_1\varepsilon_{t−27}+\theta_4\Theta_1\varepsilon_{t−28}−\Theta_2\varepsilon_{t−48}+ \theta_1\Theta_2\varepsilon_{t−49} +\theta_2\Theta_2\varepsilon_{t−50}+\theta_3\Theta_2\varepsilon_{t−51}+\theta_4\Theta_2\varepsilon_{t−52}$$




 





## ARIMA(2,1,1)(1,0,1)[24]

This notation translates into: 
$$(1−\phi_1B−\phi_2B^2)(1−\Phi_1B^24)(1−B)y_t=(1−\theta_1B)(1−\Theta_1B^24)\varepsilon_t$$

which fully expands to:


$$ y_t−y_{t−1}−\phi_1y_{t−1}+\phi_1y_{t−2}−\phi_2y_{t−2}+\phi_2y_{t−3}−\Phi_1y_{t−24}+\Phi_1y_{t−25}+\phi_1\Phi_1y_{t−25}−\phi_1\Phi_1y_{t−26}+\phi_2\Phi_1y_{t−26}−\phi_2\Phi_1y_{t−27}=\varepsilon_t−\theta_1ε_{t−1}−\Theta_1\varepsilon_{t−24}+\theta_1\Theta_1\varepsilon_{t−25}$$
